Experimental and numerical studies of the effects of resonant and random excitations of the proton beam in the LHC, with applications to the design of a pulsed hollow electron lens for active halo control

### 2018-03-03 resex.tex
Revised text of results on 10th-turn pulsing. Moved estimate of random noise to simulation section.

Quantitative predictions of loss rates and emittance growth are in
general challenging, as both observables are influenced by several
factors. For instance, the natural noise present in the machine
(originating from mechanical vibrations, current ripple in the magnet
power supplies, etc.) may have complex interactions with the external
excitations. In the LHC, noise sources at top energy are well
characterized, whereas they are not well known at injection,
presenting an influential but undefined input for simulations.

@@

\subsection{$10^{\mathrm{th}}$ turn pulsing\label{sec:simex10}}
The $10^{\mathrm{th}}$~turn pulsing pattern was tested in 2016 with an excitation in the vertical plane only. As this was the first time a resonant excitation was tested in the LHC, only in total 48~bunches could be injected in order to still guarantee the protection of the machine. To test different excitation amplitudes during one fill and have enough statistics, the excitation was therefore only applied in one plane. As described earlier, the experiment started with a period of 12~minutes without excitation in order to let the beam distribution fully adjust to its equilibrium state after injection. Afterwards the excitation was applied with the excitation scheme shown in Fig.~\ref{fig:fill} and a maximum excitation amplitude of $A_{\mathrm{max}}=5\Delta A = 48~$nrad for a total of 11~minutes. The excitation was then further increased to the maximum value reachable without saturation of $A_{\mathrm{max}}=5\Delta A = 96~$nrad and kept for 11~minutes. In both cases, the excitation induced the following changes:
\begin{enumerate}
	\item Increase of the loss rate with the excitation amplitude (Figs.~\ref{fig:10thexp}.
	\item Increase of the emittance growth with the excitation amplitude in the vertical plane but not in the horizontal plane (Fig.~\ref{fig:10thexp}).
	\item Change of the beam distribution for the excited bunches (Fig.~\ref{fig:10thexpprof}).	
\end{enumerate}
As loss rates, emittance growth and beam distribution stayed unchanged for the $4$~reference bunches, the above observations can be also truly associated with the effect of the $10^{\mathrm{th}}$~turn pulsing.
\begin{figure*}[h]
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_bunch_intensity_v10th_no_damper_avg.png}
	\end{minipage}
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emith_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emitv_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:10thexp} Relative intensity (top) and relative horizontal (bottom left) and vertical (bottom right) emittance averaged over the bunches experiencing the same excitation amplitude during the 2016 experiments and for $10^{\mathrm{th}}$~turn pulsing in~V. For all bunches the damper was not active. Relative refers here to the normalization to the initial value. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad indicated with black arrows, the excitation amplitude was linearly increased for each group of 4~bunches (see Fig.~\ref{fig:fill}). The different colored solid lines labeled with $n\Delta A$ are the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation shown as envelope. The area with a blue background indicates the fast adjustment period of the beam distribution transitioning into a new equilibrium state highlighted with a gray background.}
\end{figure*}
These experimentally obtained scalings of the loss rate and emittance growth with the excitation amplitude can be then used for the actual specification of the hollow electron lens.

Besides the numerical determination of tolerances, the vertical emittance actually features indeed the strange and interesting behavior predicted in simulations: a fast adjustment phase of the beam distribution visible as a rapid increase of the emittance followed by the reach of a new equilibrium distribution apparent by a slower continuous emittance growth. These two phases are indicated in Fig.~\ref{fig:10thexp} in blue and black.
		reference bunch, no excitation
		\includegraphics[width=1.0\linewidth]{profile_v_10thv_slot_50.png}
		excited bunch, $10^{\mathrm{th}}$~turn V		\includegraphics[width=1.0\linewidth]{profile_v_10thv_slot_1300.png}
	\caption{\label{fig:10thexpprof} Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2016 experiments. The profiles are taken at the end of the $10^{\mathrm{th}}$~turn pulsing in V. For both bunches the transverse damper is not active. The black line in the plot of the residual is the current profile minus the profile before applying the excitation and is a measure for the overall change of the distribution. The red line is the current profile minus its Gaussian fit, which gives and indication for the deviation of the distribution from a Gaussian distribution. For details on the analysis it is referred to \cite{bsrtprofinj}. The distribution of the reference bunch (left), here slot 50, stays unchanged, while the bunch experiencing the maximum excitation (right), here slot 1300, clearly shows a change to a non-Gaussian distribution.}
Thanks to the good performance of the Beam Synchrotron Radiation Telescope (BSRT)~\cite{bsrtprofinj}, the distribution changes indirectly observed as changes of the emittance could also be for the first time directly measured. As example Fig.~\ref{fig:10thexpprof} shows the vertical profile of a reference bunch and one bunch experiencing the maximum excitation of $5\cdot\Delta A=A_{\mathrm{max}}=48$~nrad and afterwards 96~nrad. The distribution of the excited bunch not only changes but becomes clearly non-Gaussian, while the reference bunch stays unchanged.
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_intensity.png}
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2_rel.png}
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit1.png}
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2.png}
	\caption{\label{fig:10thsim} Relative intensity (top left) and relative vert. emittance (top right), and hor. (bottom left) and vert. (bottom right) emittance obtained from simulations (distribution tracking) based on the 2016 injection optics with standard errors and $(Q_x,Q_y)=(64.28,59.31)$. The solid line indicates an excitation with only $10^{\mathrm{th}}$~turn pulsing in V and the dotted line with $10^{\mathrm{th}}$~turn pulsing in V plus a random dipole noise component in H+V of 6~nrad.}
The simulation results from the distribution tracking for $10^{\mathrm{th}}$~turn pulsing and without (solid line) and with (dotted line) additional random noise component to emulate the natural noise present in the LHC are shown in Fig.~\ref{fig:10thsim}. As no estimate of the natural noise is available at injection for the LHC, a first estimate is obtained by scaling the estimate for 6.5~TeV \cite{md1433_noise_top_energy,md_noise_bbLHC} by the magnetic rigidity to the injection energy of 450~GeV. This yields a maximum kick amplitude at the transverse damper of approximately
\begin{equation}
\theta_{\mathrm{random,ADT,max}}(\mathrm{450~GeV}) = 6~\mathrm{nrad}.
\end{equation}
As can be seen from simulations, adding a random noise component not only yields the correct qualitative behavior of the emittance but also increases the losses. As the losses are in general underestimated in simulations compared to the experimental results, the discrepancy between simulations and experiment can be thus reduced.  To better compare the experimental results for the two different maximum excitation amplitudes applied for slightly different time intervals and also the simulations conducted for a much shorter interval of only 90~s, we define the relative loss rate $R$:
\begin{equation}\label{eqn:lossrate}
R = \frac{I_{\mathrm{start}}-I_{\mathrm{end}}}{I_{\mathrm{start}}\cdot \Delta t}
\end{equation}
where $I$ is the intensity and $\Delta t$ the time interval during which the excitation with fixed $A_{\mathrm{max}}$ is applied during the experiments or the duration of the simulation respectively. A qualitative comparison of the loss rates obtained in experiments and simulations is shown in Fig.~\ref{fig:10thexploss} featuring a surprisingly good agreement. It should be also kept in mind that the additional random noise component considerably influences the simulation results and as long as it is unknown should be seen more as a fit parameter than a real experimentally confirmed input parameter. The measured loss rates for $A_{\mathrm{max}}=48$~nrad and $A_{\mathrm{max}}=96$~nrad should furthermore follow the same scaling. Instead the loss rates for the same excitation amplitude are smaller for $A_{\mathrm{max}}=96$~nrad compared to $A_{\mathrm{max}}=48$~nrad. That losses and also emittance growth for a fixed excitation amplitude do not exactly agree for both experiments, is a general tendency observed in all experiments. As the second excitation however starts from an already by the previous excitation changed beam distribution, the difference is however not surprising and can be explained by this difference in initial conditions. For checking the reproducibility of the results in a possible future experiment, it would be interesting to see if the same loss rates are measured if for each $A_{\mathrm{max}}$ a new fill is used. 
\begin{figure}[h]
%		\includegraphics[width=0.8\linewidth]{2016_scale_amp_10v_ran_lbllong_sim.png}
\includegraphics[width=1.0\linewidth]{2016_scale_amp_10v_ran_lblshort_sim_prstab.png}
	\caption{\label{fig:10thexploss} Comparison of scaling of loss rates $R$ as defined in Eqn.~\ref{eqn:lossrate} with the excitation amplitude for $10^{\mathrm{th}}$~turn pulsing in~V. The experimental results for a maximum excitation of $A_{\mathrm{max}}=48$~nrad are shown in black and for $A_{\mathrm{max}}=96$~nrad in green. The excitation amplitude error is 50\% and the error on the loss rate is the pure statistical error on the mean. The simulation results (see Fig.~\ref{fig:10thsim}) for 2016 injection optics, standard errors, $(Q_x,Q_y)=(64.28,59.31)$, $10^{\mathrm{th}}$~turn pulsing in V plus random dipole noise in H+V of 6~nrad are shown in red together with their statistical error. The loss rates follow approximately a quadratic behavior illustrated by a second order polynomial fit to the data shown as solid (for bunches with transverse damper off) or dashed (for bunches with transverse damper on) line.}
\end{figure}

>>>>

Quantitative predictions of loss rates and emittance growth are in
general challenging, as both observables are influenced by several
factors. For instance, the natural noise present in the machine
(originating from mechanical vibrations, current ripple in the magnet
power supplies, etc.) may have complex interactions with the external
excitations. In the LHC, noise sources at top energy are well
characterized, whereas they are not well known at injection,
presenting an influential but undefined input for simulations.  An
estimate is obtained by scaling the value at
6.5~TeV~\cite{md1433_noise_top_energy, md_noise_bbLHC} with proton
magnetic rigidity to the injection energy of 450~GeV. This yields a
maximum kick amplitude at the transverse damper of approximately
%
\begin{equation}
  \label{eq:noise-kick}
  \theta_{\mathrm{random,ADT,max}}(\mathrm{450~GeV}) = \q{6}{nrad}.
\end{equation}

@@

\subsection{Pulsing every $10^{\mathrm{th}}$ turn}
\label{sec:simex10}

\begin{figure*}
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_bunch_intensity_v10th_no_damper_avg.png}
	\end{minipage}
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emith_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emitv_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:10thexp} Relative intensity (top) and relative horizontal (bottom left) and vertical (bottom right) emittance averaged over the bunches experiencing the same excitation amplitude during the 2016 experiments and for $10^{\mathrm{th}}$~turn pulsing in~V. For all bunches the damper was not active. Relative refers here to the normalization to the initial value. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad indicated with black arrows, the excitation amplitude was linearly increased for each group of 4~bunches (see Fig.~\ref{fig:fill}). The different colored solid lines labeled with $n\Delta A$ are the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation shown as envelope. The area with a blue background indicates the fast adjustment period of the beam distribution transitioning into a new equilibrium state highlighted with a gray background.}
\end{figure*}

\begin{figure*}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		reference bunch, no excitation
		\includegraphics[width=1.0\linewidth]{profile_v_10thv_slot_50.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		excited bunch, $10^{\mathrm{th}}$~turn V		\includegraphics[width=1.0\linewidth]{profile_v_10thv_slot_1300.png}
	\end{minipage}
	\caption{\label{fig:10thexpprof} Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2016 experiments. The profiles are taken at the end of the $10^{\mathrm{th}}$~turn pulsing in V. For both bunches the transverse damper is not active. The black line in the plot of the residual is the current profile minus the profile before applying the excitation and is a measure for the overall change of the distribution. The red line is the current profile minus its Gaussian fit, which gives and indication for the deviation of the distribution from a Gaussian distribution. For details on the analysis it is referred to \cite{bsrtprofinj}. The distribution of the reference bunch (left), here slot 50, stays unchanged, while the bunch experiencing the maximum excitation (right), here slot 1300, clearly shows a change to a non-Gaussian distribution.}
\end{figure*}

\begin{figure*}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2_rel.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:10thsim} Relative intensity (top left) and relative vert. emittance (top right), and hor. (bottom left) and vert. (bottom right) emittance obtained from simulations (distribution tracking) based on the 2016 injection optics with standard errors and $(Q_x,Q_y)=(64.28,59.31)$. The solid line indicates an excitation with only $10^{\mathrm{th}}$~turn pulsing in V and the dotted line with $10^{\mathrm{th}}$~turn pulsing in V plus a random dipole noise component in H+V of 6~nrad.}
\end{figure*}

\begin{figure}
		\centering
%		\includegraphics[width=0.8\linewidth]{2016_scale_amp_10v_ran_lbllong_sim.png}
\includegraphics[width=1.0\linewidth]{2016_scale_amp_10v_ran_lblshort_sim_prstab.png}
	\caption{\label{fig:10thexploss} Comparison of scaling of loss rates $R$ as defined in Eqn.~\ref{eqn:lossrate} with the excitation amplitude for $10^{\mathrm{th}}$~turn pulsing in~V. The experimental results for a maximum excitation of $A_{\mathrm{max}}=48$~nrad are shown in black and for $A_{\mathrm{max}}=96$~nrad in green. The excitation amplitude error is 50\% and the error on the loss rate is the pure statistical error on the mean. The simulation results (see Fig.~\ref{fig:10thsim}) for 2016 injection optics, standard errors, $(Q_x,Q_y)=(64.28,59.31)$, $10^{\mathrm{th}}$~turn pulsing in V plus random dipole noise in H+V of 6~nrad are shown in red together with their statistical error. The loss rates follow approximately a quadratic behavior illustrated by a second order polynomial fit to the data shown as solid (for bunches with transverse damper off) or dashed (for bunches with transverse damper on) line.}
\end{figure}

The $10^{\mathrm{th}}$~turn pulsing pattern was tested in 2016 with an
excitation in the vertical plane only. As this was the first time a
resonant excitation was tested in the LHC, only 48~bunches were
injected, in order to guarantee safe operation of the machine. To test
different excitation amplitudes during one fill and have enough
statistics, the excitation was only applied in one plane. The
experiment started with a period of 12~minutes without excitation, to
let the beam distribution fully adjust to its equilibrium state after
injection. The excitation was then applied for 11~minutes, with the
excitation scheme shown in Fig.~\ref{fig:fill} and a maximum
excitation amplitude of
$A_{\mathrm{max}} = 5 \, \Delta A = \q{48}{nrad}$. The excitation was
then further increased to the maximum value of
$A_{\mathrm{max}} = 5 \, \Delta A = \q{96}{nrad}$ and kept for another
period of 11~minutes.

The main observations are collected in Figures~\ref{fig:10thexp}
and~\ref{fig:10thexpprof}. In Fig.~\ref{fig:10thexp}, one can see the
evolution in time of losses and emittances for the control bunches and
for the affected bunches, as a function of excitation amplitude. In
brief, during both phases of the experiment, the excitation induced
the following changes:
%
\begin{itemize}
  \item loss rates increased with excitation amplitude
    (Fig.~\ref{fig:10thexp}, left);
  \item emittance growth increased with amplitude in the vertical
    plane, but not in the horizontal plane (Fig.~\ref{fig:10thexp},
    center and right);
  \item a change of the beam distribution for the excited bunches was
    directly observed (Fig.~\ref{fig:10thexpprof}).
\end{itemize}

As loss rates, emittance-growth rates and beam distributions of the 4
reference bunches were unchanged, the above observations can be
directly attributed to the pulsing pattern. The quantitative
dependence of the loss rates and emittance-growth rates on excitation
amplitude measured in these experiments provides an estimate of the
effects of external resonant excitations, such as the residual fields
of a pulsed hollow electron lens.

The vertical emittance features a behavior similar to that predicted
in simulations: a fast adjustment phase of the beam distribution,
which manifests itself as a rapid increase of the emittance, followed
by a new equilibrium, inferred from a slower and continuous emittance
growth. In Fig.~\ref{fig:10thexp}, these two phases are indicated in
blue and black.

Distribution changes could also be directly observed, thanks to the
performance of the Beam Synchrotron Radiation Telescope
(BSRT)~\cite{bsrtprofinj}. As an example, Fig.~\ref{fig:10thexpprof}
shows the vertical profile of a reference bunch and of a bunch
experiencing the maximum excitation
$A_{\mathrm{max}} = 5 \, \Delta A = \q{48}{nrad}$ and later
96~nrad. While the reference bunch stays unchanged, the distribution
of the excited bunch clearly changes and re-adjusts to a non-Gaussian
shape, with particles shifting from the core towards the regions at
approximately $\pm 2\sigma$. This behavior is similar to what was
calculated in simulations, as discussed in Section~\ref{sec:pattern}
and in Fig.~\ref{fig:patternhist}.

Numerical simulations of intensities and emittances for this
experiment are shown in Fig.~\ref{fig:10thsim}. They are the results
of distribution tracking for $10^{\mathrm{th}}$~turn pulsing without
additional noise (solid lines) and with an additional noise component
(dotted lines), to emulate the natural noise present in the LHC
(Eq.~\ref{eq:noise-kick}). The presence of noise significantly changes
the calculated effect of the resonant excitation on both losses and
emittances.

A direct comparison of measured and predicted loss rates is presented
in Fig.~\ref{fig:10thexploss}. As experiments and simulations were
conducted on different time intervals (11~minutes vs.\ 90~s), we
compare the relative average loss rate~$R$:
%
\begin{equation}
  \label{eqn:lossrate}
  R = \frac{I_{\mathrm{start}} - I_{\mathrm{end}}}{I_{\mathrm{start}}}
  \cdot \frac{1}{\Delta t},
\end{equation}
%
where $I$ is the beam intensity and
$\Delta t = t_\mathrm{end} - t_\mathrm{start}$ is the time interval
during which the excitation with given amplitude is
applied. Uncertainties on the excitation amplitude arise from the
calibration of the transverse damping system
(Section~\ref{sec:adt}). For both experiments and simulations, the
error bars on the loss rate are statistical. The magnitude of the
random noise component introduces a systematic uncertainty in the
calculations, which was not evaluated for the scope of this paper. The
systematic uncertainty on the measurements can be inferred from the
difference between the 2~periods of the experiment, which should in
principle be the same. Part of this systematic effect could be due to
the fact that the second set of excitations with double amplitude
started with a modified beam distribution. The magnitude of this
effect could be verified in future experiments in which the
excitations are reversed or where a new beam fill is used for each set
of measurements. Overall, we conclude that the effects of this
resonant excitation could be measured in the LHC within a factor~2 and
could be predicted with similar accuracy.
